---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# discAUC

The goal of discAUC is to provide a solution to easily calculate AUC for delay
discounting data. It includes logAUC and ordAUC as published in Borges et al. 
(2016). It also includes a solution for 0 delays for logAUC.

Please cite this package using: Friedel, J. E., Treem, K. M. A., Frye, C. C. J., Salem, S. K., Westberry-Nix, M. B., & Devonshire, L. (2025). Clarifications for calculating area under the curve for discounting data: A primer and technical report. *Journal of the Experimental Analysis of Behavior*, *124*(1), e70041. [https://doi.org/10.1002/jeab.70041](https://doi.org/10.1002/jeab.70041)

## Installation

You can install the released version of discAUC from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("discAUC")
```

This is a basic example which shows you how to solve a common problem:

```{r example}
library(discAUC)

#Calculate AUC for proportional indiference points for each outcome per subject.
AUC(dat = examp_DD,
    x_axis = "delay_months",
    indiff = "prop_indiff",
    amount = 1,
    groupings = c("subject","outcome"))
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/jefriedel/discAUC/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jefriedel/discAUC/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->
